---
title: "Exploring IMDB data"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# clear memory
rm(list = ls())
gc()

# set seed for replication of randomized functions
set.seed(12345)

# set paths to data
raw_data_path = "~/Dropbox/McGill/teaching/2023-2024/summer/ORGB690/data/imdb/"

# load tidyverse
library(tidyverse, quietly = TRUE )
```

## Introduction: Goals and Strategy

In this exercise, I am going to explore the IMDB dataset. I need to...

1.  Understand the data structure and scope

2.  Identify any apparent encoding errors or points to correct

3.  Propose the research questions to consider in the course and the scope of data required for that

But before I proceed, I need to load the data. The data files were obtained from this page: <https://datasets.imdbws.com/> and the data dictionaries are available here: <https://developer.imdb.com/non-commercial-datasets/>

### Loading raw data files

```{r load-data}
# Load the required packages
library(readr) # for read_tsv() function
library(arrow) # for write_feather() function

##### read files individually and save in feather format
# names of all actors
name_basics <- read_tsv(paste0(raw_data_path, "name.basics.tsv"), na = "\\N",
                        col_types = list(
                          nconst = col_character(),
                          primaryName = col_character(),
                          birthYear = col_integer(),
                          deathYear = col_integer(),
                          primaryProfession = col_character(),
                          knownForTitles = col_character()
                        )) 

write_feather(name_basics, paste0(raw_data_path,"people.feather"))

# all titles
title_basics <- read_tsv(paste0(raw_data_path, "title.basics.tsv"), na = "\\N", quote = "",
                        col_types = list(
                          tconst = col_character(),
                          titleType = col_character(),
                          primaryTitle = col_character(),
                          originalTitle = col_character(),
                          isAdult = col_integer(),
                          startYear = col_integer(),
                          endYear = col_integer(),
                          runtimeMinutes = col_integer(),
                          genres = col_character()
                        )) 

# get a subset of just movies
movies <- title_basics |> 
  filter(titleType=="movie")

write_feather(name_basics, paste0(raw_data_path,"movies.feather"))

## principals
title_principals <- read_tsv(paste0(raw_data_path, "title.principals.tsv"), na = "\\N", quote = "", col_types = list(ordering = col_integer()))

```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
